FAIR fission track analysis with geochron@home

### 2025-08-20 geochron@home/manuscript/paper.tex
Trimmed down crowdsourcing section

%\begin{figure}[!ht]
{ \centering \includegraphics[width=\linewidth]{AvP.pdf}
  \captionof{figure}{Comparison of fission track data for Mount
    Dromedary apatite by two analysts (PV = Pieter Vermeesch and AC =
    Andrew Carter). a) the track counts and ROIs of both users for
    grain 4648, shown in blue (PV) and red (AC); b) comparison of the
    track densities for PV and AC for all 25 grains analysed by the
    two analysts with a reference line for
    $\hat{\rho}_\text{PV}=\hat{\rho}_\text{AC}$; c) radial plot of the
    same data, using Equation~\ref{eq:NO}.}
  \label{fig:AvP}
}%\end{figure}

@@

%\begin{figure}[!ht]
{ \centering \includegraphics[width=15cm]{radialcrowd.pdf}
  \captionof{figure}{a) Box plots and summary statistics of numbers of
    spontaneous fission tracks counted in each of 25 grains by 68
    novice counters (geology students) all using the same ROIs. Each
    student counted tracks in a subset of the grains. b) The track
    counts for Grain 25 plotted against those for Grain 23 for 37
    students who counted both of those grains. c) Radial plot of the
    ratios of the counts in panel b) using the precisions given by
    Equation~\ref{eq:N12}. In each panel, the counts made by PV are
    added as blue dots.\medskip}
  \label{fig:radialcrowd}
}%\end{figure}

In 1906, Sir Francis Galton visited a county fair in which a contest
was held to guess the weight of an ox. 787 people participated in the
event. Galton discovered that the median of all their estimates was
within 0.8\% of the true weight of the ox and more accurate than 90\%
of the individual estimates. Such is the `wisdom of crowds'
\citep{galton1907a}. Similar effects are seen in fission track
geochronology.\medskip

An interlaboratory comparison study by \citet{miller1985} showed that
the average of several fission track age estimates is closer to the
known age of mineral standards than the age obtained by any individual
observer. Routine measurement of fission track samples by multiple
analysts is prohibitively expensive in a normal laboratory
environment. geochron@home changes this by bringing fission track
analysis to a proverbial `county fair' of citizen scientists.\medskip

We here present the results of a type of `crowd-sourcing' experiment
that was carried out at UCL as part of an undergraduate course in
isotope geology. 70 students were asked to create an account on the
geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), the students were asked to
count fission tracks in randomly assigned image stacks of Mount
Dromedary apatite. Each student was required to analyse at least 15
grains, using pre-defined ROIs. In a matter of hours, the students
amassed a dataset of 37,331 fission track counts in 34 grains.  This
large dataset was parsed into separate data files --- one for each
student --- which they processed during an assessed practical
exercise.\medskip

The raw counts for 68 students (two of the students were unable to
complete the exercise) are given in Table~\ref{tab:crowdtable} of the
Appendix, for the same selection of 25 grains that were analysed by AC
and PV in Section~\ref{sec:private}, along with corresponding counts
made by PV. Everyone counted tracks in the same ROIs.
Figure~\ref{fig:radialcrowd}a shows box plots and summary statistics
of these counts for each grain.  Variation between the mean values is
to be expected because of the differing numbers of tracks due to
varying areas and U contents of the grains. But the variation between
counts within grains (shown by the standard deviations) is due
entirely to differences between the students’ recognising and counting
exactly the same tracks. These standard deviations increase with the
mean number of tracks and are considerable in size, being on average
about 40\% of the mean. With expert trained counters one would expect
much smaller differences between counts.  Furthermore, the vast
majority of students counted fewer tracks than PV did, often many
fewer, and on two occasions someone counted no tracks at all. PV's
count is always above the students' median and nearly always above the
upper quartile.  Table~\ref{tab:crowdtable} (looking along rows) shows
just 7 students with average counts within 10\% of PV's, and 19 with
average counts less than half of his.\medskip

Figure~\ref{fig:radialcrowd}b shows a scatter plot of pairs of counts
for two grains made by 36 students who counted both of them. They
average around 29 and 30 tracks per grain compared with PV's pair of
52 and 50, which is highlighted in blue. There is a positive
correlation between the pairs, consistent with systematic observer
effects (i.e., people who count a low value in one grain tend to count
similarly low in the other, and vice versa).  Comparing other pairs of
grains shows similar results, with correlations varying between 0.4
and 0.9. In all cases we have looked at, the residual variation, after
allowing for systematic differences between students, is consistent
with Poisson variation.  Figure~\ref{fig:radialcrowd}c shows a radial
plot of the ratios of the counts in Panel~b, using the precisions
given by
\begin{equation}
  \frac{\mbox{se}(N_1/N_2)}{N_1/N_2}
  \approx
  \sqrt{
    \frac{1}{N_1} + \frac{1}{N_2}
  }
  \label{eq:N12}
\end{equation}

The context here is rather different from that in
Figure~\ref{fig:AvP}b, where we were comparing the ratios of counts
made by two analysts on each of 25 grains. Here we are comparing the
ratios of counts for the same two grains made by each of 37
analysts. There is no theoretical justification for
Equation~\ref{eq:N12} here, but it still provides a useful benchmark
for assessing variation between the ratios. In
Figure~\ref{fig:radialcrowd}c, that variation is consistent with
Poisson variation (as it is for most other pairs of grains) but with
expert counters one would expect less variation than that. However, in
spite of the under-counting by the students, the pooled students'
ratio of 0.97 is close to PV's ratio of 1.04. This happens quite often
in this data set, but not always, and there are some pairs of grains
where the students' pooled ratio differs somewhat from PV's.
\medskip

As mentioned in Section~\ref{sec:architecture}, geochron@home stores
the actual track positions marked by the users. This raw data can be
downloaded as a \texttt{.json} file and inspected in detail.
Figure~\ref{fig:crowdsourcing} is a two-dimensional histogram for the
x-y positions of all the track positions generated by the students in
the two apatite crystals. Visual comparison of the histogram with the
optical image confirms that the students unanimously identified the
most obvious semitracks, which contain a clearly visible etch pit and
tail. Shorter and fainter tracks received fewer clicks. Datasets like
this can be used to replace integer counts of fission tracks with
probabilities, reflecting the ambiguity of some fission track
datasets. Figure~\ref{fig:crowdsourcing} also shows that some students
counted the tails of the fission tracks rather than their etch pits,
despite being told the opposite in the tutorial. Fixing this issue
will require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).\medskip

%\begin{figure}[!ht]
{  \centering
  \includegraphics[width=9cm]{4649vs4673.pdf}
  \captionof{figure}{
    a,b) Optical images in transmitted light of 
    Mount Dromedary apatites 4649 and 4673 in the crowd-sourcing experiment.
    c,d) Two-dimensional histogram of the track locations for the
    two grains, as identified by the citizen scientists.
    \medskip}
  \label{fig:crowdsourcing}
}%\end{figure}

@@

\appendix
\section{Summary tables for the crowd-sourcing experiment}

>>>>

%\begin{figure}[!ht]
{ \centering \includegraphics[width=\linewidth]{AvP.pdf}
  \captionof{figure}{Comparison of fission track data for Mount
    Dromedary apatite by two analysts (PV = Pieter Vermeesch and AC =
    Andrew Carter). a) the track counts and ROIs of both users for
    grain~3, shown in blue (PV) and red (AC); b) comparison of the
    track densities for PV and AC for all 25 grains analysed by the
    two analysts with a reference line for
    $\hat{\rho}_\text{PV}=\hat{\rho}_\text{AC}$; c) radial plot of the
    same data, using Equation~\ref{eq:NO}.}
  \label{fig:AvP}
}%\end{figure}

@@

In 1906, Sir Francis Galton visited a county fair in which a contest
was held to guess the weight of an ox. 787 people participated in the
event. Galton discovered that the median of all their estimates was
within 0.8\% of the true weight of the ox and more accurate than 90\%
of the individual estimates. Such is the `wisdom of crowds'
\citep{galton1907a}. Similar effects are seen in fission track
geochronology. An interlaboratory comparison study by
\citet{miller1985} showed that the average of several fission track
age estimates is closer to the known age of mineral standards than the
age obtained by any individual observer. Inspired by these previous
experiments, we used geochron@home to bring fission track analysis to
a proverbial `county fair' of citizen scientists.\medskip

We here present the results of a type of `crowd-sourcing' experiment
that was carried out at UCL as part of an undergraduate course in
isotope geology. 70 students were asked to create an account on the
geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), the students were asked to
count fission tracks in randomly assigned image stacks of Mount
Dromedary apatite. Each student was required to analyse at least 15
grains, using pre-defined ROIs. In a matter of hours, the students
amassed a dataset of 37,331 fission track counts in 34 grains.  This
large dataset was parsed into separate data files --- one for each
student --- which they processed during an assessed programming
exercise.\medskip

The raw counts for 68 students (two of the students were unable to
complete the exercise) are given in Table~\ref{tab:crowdtable}, for
the same selection of 25 grains that were analysed by AC and PV in
Section~\ref{sec:private}, along with corresponding counts made by
PV. Everyone counted tracks in the same ROIs. The vast majority of
students counted fewer tracks than PV did, often many fewer, and on
two occasions someone counted no tracks at all. PV's count is always
above the students' median and nearly always above the upper quartile.
Table~\ref{tab:crowdtable} (looking along rows) shows just 7 students
with average counts within 10\% of PV's, and 18 with average counts
less than half of his.\medskip

However, in spite of the under-counting by the students, the
\emph{ratios} of the track counts between different grains are
generally quite close to PV's ratios, especially when comparing grains
with similar track densities. This is apparent from
Table~\ref{tab:crowdtable} (looking along columns).
Figure~\ref{fig:23vs25}b shows a scatter plot of pairs of counts for
two grains (23 and 25) made by 36 students who counted both of
them. They average around 29 and 30 tracks per grain compared with
PV's pair of 52 and 50, which is highlighted in blue. There is a
positive correlation between the pairs, consistent with systematic
observer effects (i.e., people who count a low value in one grain tend
to count similarly low in the other, and vice versa).  Comparing other
pairs of grains shows similar results, with correlations varying
between 0.4 and 0.9. In all cases we have looked at, the residual
variation, after allowing for systematic differences between students,
is consistent with Poisson variation.

As mentioned in Section~\ref{sec:architecture}, geochron@home stores
the actual track positions marked by the users. This raw data can be
downloaded as a \texttt{.json} file and inspected in detail.
Figure~\ref{fig:23vs25}b) shows two-dimensional histograms for the x-y
positions of all the track positions generated by the students in the
two apatite crystals. Visual comparison of the histogram with the
optical image confirms that the students unanimously identified the
most obvious semitracks, which contain a clearly visible etch pit and
tail. Shorter and fainter tracks received fewer clicks. Datasets like
this can be used to replace integer counts of fission tracks with
probabilities, reflecting the ambiguity of some fission track
datasets. Figure~\ref{fig:23vs25} also shows that some students
counted the tails of the fission tracks rather than their etch pits,
despite being told the opposite in the tutorial. Fixing this issue
will require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).\medskip

%\begin{figure}[!ht]
{ \centering \includegraphics[width=9cm]{23vs25.pdf}
  \captionof{figure}{ a) The track counts for grain~25 plotted against
    those for grain~23 for 37 students who counted both of those
    grains. b) Optical images in transmitted light of Mount Dromedary
    apatites 23 and 25 in the crowd-sourcing experiment.  c,d)
    Two-dimensional histogram of the track locations for the two
    grains, as identified by the citizen scientists.
    \medskip}
  \label{fig:23vs25}
}%\end{figure}

@@
